---
title: "Historical_Events_Timeseries"
author: "sundar pradeep"
date: "March 31, 2015"
output: html_document
---

#### Synopsis: 

This is an analysis of major disasters in human history with the following objectives: <br>

* Understand the nature of disasters.<br>
* Explore if the pattern of disasters have changed over time - with our ability to predict natural disasters and prevent diseases.<br>
* Estimate world population trends if these events had not occurred - and what share of these disasters were avoidable (man-made).<br>

#### Scope, Variables and Datasets: 

Categories of major events considered for the analysis: 

* [`Natural disasters`](http://en.wikipedia.org/wiki/List_of_natural_disasters_by_death_toll) 
* [`Man-made disasters`](http://en.wikipedia.org/wiki/List_of_wars_and_anthropogenic_disasters_by_death_toll) - wars and genocides
* [`Epidemics`](http://en.wikipedia.org/wiki/List_of_epidemics)

All recorded event of above categories with at least 10,000 casualties was considered.

While the original intention was to analyse the trends over last 2000 years , inconsistent data availability limited the scope to a shorter time frame - from 1500 AD to present.  

#### Pre-Processing:
<br>
* All datasets were sourced from Wikipedia  using readHTMLtable function. <br>
* Events were filtered for date > 1500 AD and toll > 10,000.<br>
* Location field was cleaned and geo-coded for visualization.<br>
* [Population](http://www.nber.org/papers/w14448) over time table was loaded for benchmarking.<br>


```{r, echo=FALSE, message = FALSE, warning = FALSE }
# set-up environment 

setwd("~/dev/Projects/2_timeseries_visualization/")

library("dplyr")
library("tidyr")
library("ggplot2")
library("xlsx")
library("ggthemes")
library("gridExtra")
library("XML")
library("httr")
library("vcd")
library("splines")

# load files.
source("./1_get_data.R")

# clean and generate tidy data for analysis.
source("./2.1_process_manmade_data.R")
source("./2.2_process_natural_data.R")
source("./2.3_process_epidemic_data.R")
source("./2.4_geocode_input_data.R")
source("./2.5_process_population_data.R")


```


### The big picture.

#### Disasters over the last 500 years across the globe. 

```{r, echo=FALSE, message = FALSE, warning = FALSE }
world.map <- map_data('world')

ggplot() + 
    geom_polygon(dat=world.map, aes(long, lat, group=group), fill="grey50") + 
    geom_point(data=consolidated.final, 
               aes(x=consolidated.final$lon, y=consolidated.final$lat,
                   size=log(consolidated.in$toll), color=consolidated.final$major,
                   position="jitter"))

```

<br>

* Europe, Africa and North America seems to have a disproportionate share of man-made disasters and epidemics 
* Asia seems to be most impacted by natural disasters , followed by man-made disasters.
* There has been no major epidemics in South America.

[click](http://cdb.io/1GJRTxA) for timeseries visualization.


Lets dig a little deeper.

```{r, echo=FALSE, message = FALSE, warning = FALSE }

minor.plot <- qplot(reorder(minor,toll),toll,data=consolidated.final,
                    geom="boxplot",color=major, xlab="Disaster Category", main="Casualty by Disaster Type",ylab="Death toll") + theme_bw()

time.plot <- qplot(decade,toll,data=consolidated.final,geom="area",
                   color=major,fill=major, xlab="Century", main="Casualty by Disaster Type",ylab="Death toll") + theme_bw()

grid.arrange(minor.plot, time.plot, ncol=1)

```

It is clear from Boxplot above that man-made disasters account for most of the casualties over the last 500 years. <br>
The sharp spike in the death toll over time from mid-18th century might be misleading , as the graph above considers absolute numbers and ignores the sharp increase in population over the time in consideration. <br>

Share of disaster casualties as a factor of overall population ( or current population equivalent) would be a better measure for trend analysis. Clearly, Population growth trends should be included as a factor in the analysis.

### Population: Regression modeling.

The problem with analyzing event casualties from 1500 AD as a share of total population is that population trend by decade, as required for the analysis is not available.<br> 

```{r, echo=FALSE, message = FALSE, warning = FALSE }

for (i in 1:nrow(consolidated.final)) {
    if (nrow(population.in[population.in$decade==consolidated.final$decade[i],]) >= 1 ) {
        consolidated.final$census.flag[i] = 1
    }
    else {
        consolidated.final$census.flag[i] = 0
    }
}

hist(consolidated.final$decade,breaks=51,col=consolidated.final$census.flag,include.lowest=T) 


```

Hence , I built a model to predict population form available sparse population metrics, using regression.


```{r, echo=FALSE, message = FALSE, warning = FALSE }
#population.lm <- lm(total ~ poly(date,5),data=population.in)

#qplot(date, total, data = population.in) +
#geom_smooth(method = lm, formula = y ~ ns(x, df = 4)) 

population.spline <- lm(total ~ ns(date, df = 4),
                        data = population.in)


prediction.by.decade <- predict(population.spline,decades)
decades$prediction <- prediction.by.decade

actual.plot <- qplot(date,total,data=population.in,main="actual",
                     label(x="Date",y="Population")) + 
  theme_bw()
#geom_line(data=population.in,aes(x=date,y=total),color="green") +
prediction.plot <- qplot(date,total,data=population.in,main="model") + 
    geom_line(data=decades,aes(x=date,y=prediction),color="red") + theme_bw()

overlay.plot <- qplot(date,prediction,data=decades,main="prediction",) + 
    geom_line(data=decades,aes(x=date,y=prediction),color="blue") + theme_bw()

grid.arrange(actual.plot, prediction.plot, overlay.plot, ncol=3)

```

<br>
Using this population prediction model , I calculated the growth factor for each decade and projected the current day equivalent of the casualties and replotted the disaster trend over time. 


```{r, echo=FALSE, message = FALSE, warning = FALSE }

current.population <- 7304813
decades$factor <- current.population / decades$prediction
projected <- merge(consolidated.final,decades,by="decade",all.x=T)
projected <- mutate(projected,
                    projected.toll = toll * factor)


qplot(decade,projected.toll,data=projected,
                   color=major,fill=major, xlab="Century", main="Adjusted Casualty by Disaster Type",ylab="Death toll",stat="identity") +
 geom_bar(stat = "identity") + theme_bw() + facet_wrap(~major)

```


We can see that 
* All three types show increase in occurrence from 1800 AD - it is possible that this is observed due to better record keeping. 
* There is a significant increase in man-made disasters from late 1800s , after a relatively peaceful period of almost a century.
* Most of the casualties caused by epidemics is contributed by two major episodes ( early 20th century Influenza and late 20th century AIDS)
* Man made disasters have contributed to most of the casualties , followed by epidemics , followed by natural disasters.


<br>
### What-if?

Now that we have a model to predict population and a growth factor for decade of the event in consideration, we can answer an interesting question .. what is the current day equivalent of the casualties - in other words , what would be our total world population (7.3B as-of-date) in case these disasters had not happened. 


```{r, echo=FALSE, message = FALSE, warning = FALSE }


whatif1.plot <- qplot(major,projected.toll,data=projected,
                   color=major,fill=major, xlab="Disaster Type", main="Adjusted Casualty by Disaster Type",ylab="Death toll",stat="identity") +
 geom_bar(stat = "identity") + theme_bw() 

whatif2.plot <- qplot(location,projected.toll,data=projected,
                   color=major,fill=major, xlab="Disaster Type", main="Adjusted Casualty by Country",ylab="Death toll",stat="identity") +
 geom_bar(stat = "identity") + theme_bw() 

grid.arrange(whatif1.plot, whatif2.plot, ncol=1)

#sum(projected$projected.toll)
```


### Caveats: 
* Population trends and projections are based on global averages , have not considered variations in growth trends across continents.

### Conclusion: 

While it is widely acknowledged that war have resulted in exponential advance in science and technology, we have no model at our disposal to predict how the world would have looked today if we had not lost a population equivalent to almost 2B people due to wars and man-made disasters. It will always be a guessing game. 

### Reference:

Github [Source Code](https://github.com/sundarpradeep/1_EDA_NYC_Schools)

<br>
<br>

